---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidycensus)
library(tidyverse)
library(choroplethr)
library(choroplethrMaps)
library(ggplot2)
library(dplyr)
library(tmap)
library(sf)
library(jsonlite)
library(janitor)
```


# try to map dallas zoning layers downloaded from city website... 

```{r}
dallas_zones <- st_read("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/Dallas_Base_Zoning/Base_Zoning.shp") %>% st_transform(., 4326)
```
```{r}
dallas_zones

qtm(dallas_zones)
```

#Load boundary datasets for Census Places and Texas Census Tracts 
```{r}
places <- st_read("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/tl_2024_48_place/tl_2024_48_place.shp") %>% st_transform(., 4326)

texas_tracts <- st_read("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/tract_shape_48/tl_2020_48_tract20.shp") %>%
  st_transform(., 4326)

```
Load in the census data from 2010 and 2019 
```{r}
#census_tracts_data_2010_2019 <- read_csv("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/save_census_tracts_2010_2019.csv")

#join with tracts from 2010....  

census_tracts <- read_csv("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/housing_tract_census_tx_2020.csv")

census_tracts <- census_tracts %>%
  mutate(total_units_2010 = total_units_2020 - units_built_since_2010)

census_tracts$GEOID <- as.character(census_tracts$GEOID)


#join the census tract data with the texas_tracts dataset (geometry)

 texas_tracts <- texas_tracts %>%
  left_join(., census_tracts, by = c("GEOID20" = "GEOID"))
```


# Todo next: add colors from zoning district types onto Dallas map....
```{r}

city <- "Dallas" # "Fort Worth" #  #"Austin" #"Houston" # "Dallas" # Fort Worth

city_boundary  <- places %>% filter(., NAME == city)

tracts_intersect_city <- texas_tracts[st_intersects(texas_tracts, city_boundary) %>% lengths > 0,]

#add a 100km buffer
boundary_with_buffer <- st_buffer(city_boundary$geometry, 10000)

tracts_intersect_buffer <- texas_tracts[st_intersects(texas_tracts, boundary_with_buffer) %>% lengths > 0,]

```

#display how the boundaries compare to the census tracts...
```{r}
ggplot()  +  geom_sf(data = tracts_intersect_city,  fill = NA, aes(color = "Census Tracts", geometry = geometry), color = "grey") + 
    geom_sf(data = city_boundary,  alpha = 0.1, fill = "red", aes(color = NAME,  geometry = geometry)) + 
  theme_void() +
    labs(
    title = paste(city, "City Boundaries and Census Tracts"),
#    subtitle = "ACS 2023",
    caption = "Data: US Census 2020",
    color = "City Outline")
```

# same map as above but no legend
```{r}
ggplot()  +  geom_sf(data = tracts_intersect_city, color= "grey", fill = NA, aes(color = "Census Tracts", geometry = geometry)) + 
    geom_sf(data = city_boundary,  alpha = 0.1, color = "red", fill = "red", aes(color = NAME,  geometry = geometry)) + 
  theme_void() +
    labs(
    title = paste(city, "City Boundaries and Census Tracts"),
#    subtitle = "ACS 2023",
    caption = "Data: US Census 2020",
    color = "City Outline")
```

#Load in NRI data
```{r}
NRI_tract <- read_csv("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/National_Risk_Index_Census_Tracts_-2688098782675405028.csv")

#cut it down to texas
NRI_tract <- NRI_tract %>% clean_names()

#includes data for 6883 tracts out of 6896 tracts
NRI_tx_tract <- NRI_tract %>% 
  filter(., state_fips_code == 48) %>% 
  mutate(new_rating_category = case_when(
        national_risk_index_rating_composite == "Very High" ~ "1 - Very High" , 
    national_risk_index_rating_composite == "Relatively High" ~ "2 - Relatively High" , 
    national_risk_index_rating_composite == "Relatively Moderate" ~ "3 - Relatively Moderate" , 
    national_risk_index_rating_composite == "Relatively Low" ~ "4 - Relatively Low" , 
    national_risk_index_rating_composite == "Very Low" ~ "5 - Very Low" , 
    TRUE ~ national_risk_index_rating_composite))
 
NRI_tx_tract_small <-  NRI_tx_tract %>% select(., c("county_name", "census_tract", "census_tract_fips_code", "population_2020", "building_value", "area_sq_mi","national_risk_index_value_composite", "national_risk_index_score_composite", "national_risk_index_rating_composite", "new_rating_category"))


# NRI_tx_tract_small <- NRI_tx_tract_small %>%
#   mutate(new_category = case_when(
#         national_risk_index_rating_composite == "Very High" ~ "1 - Very High" , 
#     national_risk_index_rating_composite == "Relatively High" ~ "2 - Relatively High" , 
#     national_risk_index_rating_composite == "Relatively Moderate" ~ "3 - Relatively Moderate" , 
#     national_risk_index_rating_composite == "Relatively Low" ~ "4 - Relatively Low" , 
#     national_risk_index_rating_composite == "Very Low" ~ "5 - Very Low" , 
#     TRUE ~ national_risk_index_rating_composite))

```

```{r}
# join the tracts with the NRI data....

tracts_intersect_city <- tracts_intersect_city %>%
  left_join(., NRI_tx_tract_small, by = c("GEOID20" = "census_tract_fips_code"))
```

# Try texas as a whole
```{r}
texas_tracts_plot <- texas_tracts %>%
  left_join(., NRI_tx_tract_small, by = c("GEOID20" = "census_tract_fips_code"))

ggplot()  +  geom_sf(data = texas_tracts_plot, color = NA, alpha = 0.9, aes(fill = new_rating_category, geometry = geometry)) +
 # geom_sf(data = city_boundary,  color= "red", alpha = 0.1, aes( geometry = geometry), ) + 
  theme_classic() +
  scale_fill_brewer(palette="YlGn" , direction = -1) + 
  labs(
    title = paste("Texas Census Tracts with National Risk Index"),
#    subtitle = "ACS 2023",
    caption = "Data: US Census | NRI FEMA",
    fill = "NRI Rating Composite")
```



Map it! NRI Categories
```{r}
ggplot()  +  geom_sf(data = tracts_intersect_city, alpha = 0.9, aes(fill = new_rating_category, geometry = geometry)) +
  geom_sf(data = city_boundary,  color= "red", alpha = 0.1, aes( geometry = geometry), ) + 
  theme_classic() +
  scale_fill_brewer(palette="YlGn" , direction = -1) + 
  labs(
    title = paste(city, "Census Tracts with NRI"),
#    subtitle = "ACS 2023",
    caption = "Data: US Census | NRI FEMA",
    fill = "NRI Value Composite")
```


Map the NRI with the 100km census tracts buffer...
```{r}
tracts_intersect_buffer <- tracts_intersect_buffer %>%
  left_join(., NRI_tx_tract_small, by = c("GEOID20" = "census_tract_fips_code"))
```

### Absolute RISK SCORE
```{r}
#issue: need to to add the city boundaries to the legend... 
ggplot()  +  geom_sf(data = tracts_intersect_city, aes(fill = national_risk_index_score_composite, geometry = geometry)) +
 geom_sf(data = city_boundary,  color= "red", fill= NA, aes( geometry = geometry)) +
  theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1) + 
  labs(
    title = paste(city, "Census Tracts with NRI"),
    caption = "Data: US Census | NRI FEMA", 
    fill = "NRI Risk Score")
```

# RATING CATEGORY with buffer
```{r}
ggplot()  +  geom_sf(data = tracts_intersect_buffer, aes(fill = national_risk_index_rating_composite, geometry = geometry)) +
 geom_sf(data = city_boundary,  color= "red",  alpha = 0.1, aes( geometry = geometry)) +
  theme_classic() +
  scale_fill_brewer(palette="YlGn" , direction = -1) + 
  labs(
    title = paste(city, "Census Tracts with NRI"),
    subtitle = "ACS 2023",
    caption = "Data: US Census | NRI FEMA", 
    fill = "NRI Rating")
```


#calculate the intersection of different risk categories...
What percentage of the city is in the category (national_risk_index_rating_composite)? breakdown by area... 
# break down by population? 
  -  2000 versus 2020 
```{r}

#aim: get the percentage of overlap so that we can calculate the NRI of the city... 
city_boundary <- city_boundary %>%
  mutate(area = st_area(.))

tracts_intersect_city <- st_make_valid(tracts_intersect_city)
city_boundary <- st_make_valid(city_boundary)

tracts_intersect_city_short <- tracts_intersect_city %>%   # only select columns needed to merge
  dplyr::select(national_risk_index_score_composite, national_risk_index_rating_composite, population_2020, building_value, total_units_2010, total_units_2020,  units_built_percent_since_2010, new_rating_category, geometry)

#find the intersection of the tracts and the city boundary... 
intersection <- st_intersection(tracts_intersect_city_short, city_boundary) 

intersect_pc <- intersection %>%
  mutate(intersect_area = st_area(.))

total_population <- sum(intersect_pc$population_2020)
total_land_area <- city_boundary$area
total_building_value <- sum(intersect_pc$building_value)

 
 ###final
  intersect_pc <- intersect_pc %>%
   group_by(national_risk_index_rating_composite) %>%
   summarise(total_area = sum(intersect_area), total_pop = sum(population_2020), total_bv = sum(building_value)) %>%
    mutate(pct_area = total_area*100/total_land_area, pct_pop = total_pop*100/total_population, 
           pct_building_value = total_bv*100/total_building_value)


```


```{r}
#print out the population, building value, and area by the NRI category... 
intersect_pc
```

# CLEANER -- Try mapping the  intersection so that we only get the outlines of the 
```{r}
ggplot()  +  geom_sf(data = intersection, aes(fill = new_rating_category, geometry = geometry)) +
 #geom_sf(data = city_boundary,  color= "red",  alpha = 0.1, aes( geometry = geometry)) +
  theme_classic() +
  scale_fill_brewer(palette="YlGn" , direction = -1) + 
  labs(
    title = paste(city, "Census Tracts with NRI"),
    subtitle = "ACS 2023",
    caption = "Data: US Census | NRI FEMA", 
    fill = "NRI Rating")
```

Show the population distribution (normalised by area... )
```{r}
#calculate population density by sq km 
intersection <- intersection %>%
  mutate(area_calc = st_area(.)/1000000) %>%
  mutate(pop20_normal = population_2020/area_calc) %>%
  mutate(pop20_density_no_unit = units::drop_units(pop20_normal))


#consider the z score for population density... 
ggplot()  +  geom_sf(data = intersection, aes(fill = pop20_density_no_unit, geometry = geometry)) +
 #geom_sf(data = city_boundary,  color= "red",  alpha = 0.1, aes( geometry = geometry)) +
  theme_classic() +
  scale_fill_distiller(palette="YlGn", direction = 1 , limits = c(0, 25000)) + 
  # limits = c(min(intersection$pop20_density_no_unit), max(intersection$pop20_density_no_unit))) +
  labs(
    title = paste(city, "Distribution of Population Density by Census Tracts"),
    subtitle = "US Census 2020, Density by Meter squared",
    caption = "Data: US Census", 
    fill = "Population Density")


```

```{r}
#analyse housing production 
intersection <- intersection %>%
  mutate(units_2010_normalised_nounit  = units::drop_units(total_units_2010/area_calc)) %>%
  mutate(units_2020_normalised_nounit = units::drop_units(total_units_2020/area_calc))


ggplot()  +  geom_sf(data = intersection, aes(fill = units_2010_normalised_nounit, geometry = geometry)) +
 #geom_sf(data = city_boundary,  color= "red",  alpha = 0.1, aes( geometry = geometry)) +
  theme_classic() +
  scale_fill_distiller(palette="YlGn", direction = 1,  limits = c(0,80000)) +
  labs(
    title = paste(city, "Housing Density by Census Tracts"),
    subtitle = "US Census 2010, Housing Units by Square Kilometer",
    caption = "Data: US Census", 
    fill = "Housing Density")
```

```{r}
ggplot()  +  geom_sf(data = intersection, aes(fill = units_2020_normalised_nounit, geometry = geometry)) +
 #geom_sf(data = city_boundary,  color= "red",  alpha = 0.1, aes( geometry = geometry)) +
  theme_classic() +
  scale_fill_distiller(palette="YlGn", direction = 1,  limits = c(0,8000)) +
  labs(
    title = paste(city, "Housing Density by Census Tracts"),
    subtitle = "US Census 2020, Housing Units by Square Kilometer",
    caption = "Data: US Census", 
    fill = "Housing Density")
```
```{r}
 
ggplot()  +  geom_sf(data = intersection, aes(fill = units_built_percent_since_2010, geometry = geometry)) +
 #geom_sf(data = city_boundary,  color= "red",  alpha = 0.1, aes( geometry = geometry)) +
  theme_classic() +
  scale_fill_distiller(palette="Greens", direction = 1,  limits = c(0,300)) +
  labs(
    title = paste(city, "Housing Growth"),
    subtitle = "Census Tracts Percent Change Housing Units 2010 to 2020",
    caption = "Data: US Census", 
    fill = "Housing Percentage Change")

```

```{r}
# calculate the growth of housing in which areas? 

totalunits2010 <- sum(intersection$total_units_2010)
totalunits2020 <- sum(intersection$total_units_2020)

  intersect_housing <- intersection %>%
  mutate(intersect_area = st_area(.)) %>%
   group_by(national_risk_index_rating_composite) %>%
   summarise(total_area = sum(intersect_area), total_pop = sum(population_2020), 
             total_2010hunit = sum(total_units_2010), total_2020hunit = sum(total_units_2020)) %>%
    mutate(pct_area = total_area*100/total_land_area, pct_pop = total_pop*100/total_population, 
           pct_housing10 = total_2010hunit*100 /totalunits2010, pct_housing20 = total_2020hunit*100/ totalunits2020)
  
  intersect_housing <- intersect_housing %>%
    mutate(unit_growth = total_2020hunit -total_2010hunit)

  intersect_housing
```


###DALLAS Zoning

#lets create a new variable on the dallas zoning dataset to categorize the zoning into residential or mulit
#family... 

From the Dallas Zones website: https://weborigin1.dallascityhall.com/zoning/html/residential_districts.html 

Residential Zones: 

A(A) - Agricultural District
R-1ac(A) - Single Family 1 Acre District
R-1/2ac(A) - Single Family 1/2 Acre District
R-16(A) - Single Family 16,000 Square Feet District
R-13(A) - Single Family 13,000 Square Feet District
R-10(A) - Single Family 10,000 Square Feet District
R-7.5(A) - Single Family 7,500 Square Feet District
R-5(A) - Single Family 5000 Square Feet District
D - Duplex District
TH-1(A) - Townhouse 1 District
TH-2(A) - Townhouse 2 District
TH-3(A) - Townhouse 3 District
CH - Clustered Housing District
MF-1(A) and MF-1(SAH) - Multifamily District
MF-2(A) and MF-2(SAH) - Multifamily District
MF-3(A) - Multifamily District
MF-4(A) - Multifamily District
MH(A) - Mobile Home District

Example code: 
starwars %>%
  select(name:mass, gender, species) %>%
  mutate(
    type = case_when(
      height > 200 | mass > 200 ~ "large",
      species == "Droid"        ~ "robot",
      TRUE                      ~ "other"
    )
  )

match_vector <- c("B", "F")

df_new <- df %>% 
  group_by(ID) %>% 
  mutate(Final = if_else(any(Name %in% match_vector), 
                         paste(match_vector[match_vector %in% Name], collapse = ";"), 
                         "No"))

```{r}
#mutate the original dataset... 
# add residential versus other type, and then multi family versus other 
resi_zones_vector <- c("A(A)", "R-1ac(A)", "R-1/2ac(A)", "R-16(A)", "R-13(A)", "R-10(A)",
                       "R-7.5(A)", "R-5(A)", "D", "TH-1(A)", "TH-2(A)", "TH-3(A)", "CH",
                       "MF-1(A)","MF-1(SAH)", "MF-2(A)", "MF-2(SAH)", "MF-3(A)", "MF-4(A)", 
                       "MH(A)")

multi_unit_zones_vector <- c("D", "MF-1(A)", "MF-1(SAH)", "MF-2(A)", "MF-2(SAH)", "MF-3(A)", 
                             "MF-4(A)", "MH(A)")

dallas_zones_new <- dallas_zones %>%
  group_by(ZONE_DIST) %>%
  mutate(resi_district_type = if_else(any(ZONE_DIST %in% resi_zones_vector), 
                                 TRUE,
                                 FALSE)) %>%
  mutate(multi_unit = if_else(any(ZONE_DIST %in% multi_unit_zones_vector), 
                                 TRUE,
                                 FALSE))
                                 

```

Lets map some
```{r}
ggplot()  +  geom_sf(data = dallas_zones_new, color = NA, aes(fill = resi_district_type, geometry = geometry)) +
# geom_sf(data = city_boundary,  color= "red",  alpha = 0.1, aes( geometry = geometry)) +
  theme_classic() +
  scale_fill_brewer(palette="BlGn" , direction = 1) + 
  labs(
    title = paste(city, "Residential Zoning"),
    subtitle = "ACS 2023",
    caption = "Data: US Census | NRI FEMA", 
    fill = "Residential Zoning")
```
# Calculate the values... 
```{r}
dallas_zones_new <- st_make_valid(dallas_zones_new)
#city_boundary <- st_make_valid(city_boundary)

dallas_zones_new_short <- dallas_zones_new %>%   # only select columns needed to merge
  dplyr::select(ZONE_DIST, resi_district_type, multi_unit, geometry)

intersect_dallas <- st_intersection(dallas_zones_new_short, city_boundary) 

intersect_dallas <- intersect_dallas %>%
  mutate(intersect_area = st_area(.))

city_boundary <- city_boundary %>%
  mutate(area = st_area(.))
city_boundary$area #863242278

#calculate
dallas_total_land_area <- city_boundary$area
```

# Calculate the percentages of city that is residential / non resi zoning
```{r}
# print out the portions of area... for dallas zoning
 ###final
  intersect_dallas_resi <- intersect_dallas %>%
   group_by(resi_district_type) %>%
   summarise(total_area = sum(intersect_area)) %>% #, total_pop = sum(population_2020), total_bv = sum(building_value))%>%
    mutate(pct_area = total_area*100/dallas_total_land_area) #, pct_pop = total_pop*100/total_population, 
          # pct_building_value = total_bv*100/total_building_value) 
intersect_dallas_resi
```


#Calculate the percentages of city that is multi -unit or not... 
```{r}
  ###final
  intersect_dallas_multi <- intersect_dallas %>%
   group_by(multi_unit) %>%
   summarise(total_area = sum(intersect_area)) %>% #, total_pop = sum(population_2020), total_bv = sum(building_value))%>%
    mutate(pct_area = total_area*100/dallas_total_land_area) #, pct_pop = total_pop*100/total_population, 
          # pct_building_value = total_bv*100/total_building_value)

intersect_dallas_multi

```


# map out the multi unit portions of the city....
```{r}

ggplot()  +  geom_sf(data = dallas_zones_new, color = NA, aes(fill = multi_unit, geometry = geometry)) +
# geom_sf(data = city_boundary,  color= "red",  alpha = 0.1, aes( geometry = geometry)) +
  theme_classic() +
  scale_fill_brewer(palette="YlGn" , direction = 1) + 
  labs(
    title = paste(city, "Dallas Multi Unit Zoning"),
    subtitle = "ACS 2023",
    caption = "Data: US Census | NRI FEMA", 
    fill = "Multi residential Zoning")
```

# TODO: Now overlap the zoning with the NRI... 
```{r}
dallas_zones_new_short <- st_make_valid(dallas_zones_new_short)

#dallas_zones_new_short <- dallas_zones_new_short %>%
 # mutate(area_calc = st_area(.))

#try combining the areas first... 
# combine based on national_risk_rating_composite
tracts_intersect_city_short_NRI <- tracts_intersect_city_short %>%
  select(c("new_rating_category", "geometry")) %>%
  group_by(new_rating_category) %>%
   summarise(geometry = st_union(geometry ))

#, "total_units_2010", "total_units_2020",
#%>%
#  group_by(national_risk_index_rating_composite) %>%
#  summarize(geometry= st_union(geometry), total_2010 = sum(total_units_2010), 
#            total_2020 = sum(total_units_2020))

#combine dallas zones based on zone district 
#dallas_zones_new_short_combined <- dallas_zones_new_short %>%
#  group_by(resi_district_type, multi_unit) %>%
#  summarize(geometry = st_union(geometry))

#just combine by district
dallas_zones_district <-  dallas_zones_new_short %>%
  group_by(ZONE_DIST, resi_district_type, multi_unit) %>%
  summarize(geometry = st_union(geometry))

tracts_intersect_city_short_NRI <- st_make_valid(tracts_intersect_city_short_NRI)

#dallas_zones_new_short_combined <- st_make_valid(dallas_zones_new_short_combined)

dallas_zones_district <- st_make_valid(dallas_zones_district)

 
#intersect the zoning with the NRI census tracts 
#NRI_zoning_dallas <- st_intersection(tracts_intersect_city_short_combined, dallas_zones_new_short_combined) 

NRI_zoning_dallas <- st_intersection(tracts_intersect_city_short_NRI, dallas_zones_district, model = "closed") 

dallas_zones_district <- st_make_valid(dallas_zones_district)

sf_use_s2(FALSE)


NRI_zoning_dallas <- NRI_zoning_dallas %>%
  mutate(area_calc = st_area(.))

#NRI_zoning_dallas_grouped <- NRI_zoning_dallas %>%
#  group_by(ZONE_DIST, national_risk_index_rating_composite) %>%
#  summarize(geometry = st_union(geometry)) 

NRI_zoning_dallas_no_geom <- NRI_zoning_dallas %>%
  st_drop_geometry()


NRI_zoning_dallas_no_geom_grouped <- NRI_zoning_dallas_no_geom %>%
  select(-c(ZONE_DIST)) %>%
  group_by(new_rating_category, resi_district_type, multi_unit) %>%
  summarize(total_area = sum(area_calc))


#calculate
total_zoning_land_area <- sum(NRI_zoning_dallas_no_geom_grouped$total_area)

dallas_total_land_area <- city_boundary$area


NRI_zoning_dallas_no_geom_grouped <- NRI_zoning_dallas_no_geom_grouped %>%
  mutate(pct_area = total_area*100/total_zoning_land_area, total_area_km = total_area/1000000)
```

```{r}
NRI_zoning_dallas_resi <- NRI_zoning_dallas %>%
  filter(resi_district_type == TRUE)

```

```{r}
ggplot()  +  geom_sf(data = NRI_zoning_dallas_resi, color = NA, aes(fill = new_rating_category, geometry = geometry)) +
# geom_sf(data = city_boundary,  color= "red",  alpha = 0.1, aes( geometry = geometry)) +
  theme_classic() +
  scale_fill_brewer(palette="YlGn" , direction = -1) + 
  labs(
    title = paste(city, "Residential Zoning by NRI"),
    subtitle = "ACS 2020",
    caption = "Data: US Census | NRI FEMA", 
    fill = "NRI Rating")
```


```{r}

NRI_zoning_dallas_multi <- NRI_zoning_dallas %>%
  filter(resi_district_type == TRUE & multi_unit == TRUE)

ggplot()  +  geom_sf(data = NRI_zoning_dallas_multi, color = NA, aes(fill = new_rating_category, geometry = geometry)) +
# geom_sf(data = city_boundary,  color= "red",  alpha = 0.1, aes( geometry = geometry)) +
  theme_classic() +
  scale_fill_brewer(palette="YlGn" , direction = -1) + 
  labs(
    title = paste(city, "Residential Multi-family Zoning by NRI"),
    subtitle = "ACS 2020",
    caption = "Data: US Census | NRI FEMA", 
    fill = "NRI Rating")
```

### San antonio 
```{r}

SA_Zoning <- read_csv("Data/San_Antonio_Zoning.csv")
```

