---
title: "R Notebook"
output: html_notebook
---
script to generate two data files. 
Output: 
- 'housing_tract_census_tx.csv'
- 'save_census_data_tract.csv' 


```{r}
library(tidycensus)
library(tidyverse)
library(choroplethr)
library(choroplethrMaps)
library(ggplot2)
library(dplyr)
library(tmap)
library(sf)
```

#Dataset 1  - Census Tracts 2023 - units and year built 
```{r}
#Let's try querying a few more moving variables to estimate how many people have moved...?
#B07013_013E - has to do with moved in last year...
#
#B25034_001E total number of units I think ? 
#B25034_002 2020 or later
#B25034_003 2010 to 2019 

tract_tx_new20 <- get_acs(geography = "tract",
                 variables = c("B25034_002"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2020)

tract_tx_new10 <- get_acs(geography = "tract",
                 variables = c("B25034_003"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2020)

tract_tx_total_hunit <- get_acs(geography = "tract",
                 variables = c("B25034_001"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2020)

```
Lets join these together.... 
```{r}
#slim down the datasets to remove the columns we don't need
tract_tx_new10 <- tract_tx_new10 %>% 
  select(c("GEOID", "estimate")) %>%
  rename(units_built_2010_19 = estimate)

tract_tx_new20 <- tract_tx_new20 %>%
  select(c("GEOID", "estimate")) %>%
  rename(units_built_2020 = estimate)

tract_tx_total_hunit <- tract_tx_total_hunit %>%
  rename(total_units_2020 = estimate) %>%
  left_join(., tract_tx_new10, by = c("GEOID" = "GEOID")) %>%
  left_join(., tract_tx_new20, by = c("GEOID" = "GEOID") )

tract_tx_total_hunit <- tract_tx_total_hunit %>%
  mutate(units_built_since_2010 = (units_built_2010_19 + units_built_2020)) %>%
  mutate(units_built_percent_since_2010 = (100*units_built_since_2010 / (total_units_2020 -units_built_since_2010)))
  

head(tract_tx_total_hunit)
```

```{r}
write.csv(tract_tx_total_hunit, "~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/housing_tract_census_tx_2020.csv") 
```


#Dataset 2 
Also get the data for population, etc... 

## FETCH the data at Texas Tract levels: Population, Housing units, Median Home Value

```{r}
# DP05_0001 Total population 
# DP05_0091 number of housing units 
# B25077_001 Median home value
# B07101_001E is the number of movers from another region... 
#tract_tx_pop <- 

#population tract  
# tract_tx_pop <- get_acs(geography = "tract",
#                  variables = c("DP05_0001"),
#                 state = "TX",
#               #  county = 201,
#                 survey = "acs5",
#                 year = 2020)
# 
# tract_tx_hunit <- get_acs(geography = "tract",
#                  variables = c("DP05_0091"),
#                 state = "TX",
#               #  county = 201,
#                 survey = "acs5",
#                 year = 2020)


#but use the decennial not ACS 

tract_tx_pop_2020 <- get_decennial(geography = "tract",
                 variables = c("P1_001N"),
                state = "TX",
                year = 2020)

tract_tx_hunit_2020 <- get_decennial(geography = "tract",
                 variables = c("H1_001N"),
                state = "TX",
                year = 2020)
```


### Manipulate the data  and merge into one dataset
```{r}
# drop unneeded columns
tract_tx_pop_2020 <- tract_tx_pop_2020 %>% select(., c("GEOID", value )) %>%
  rename(., population20 = value) 

tract_tx_hunit_2020 <- tract_tx_hunit_2020 %>% select(., c("GEOID", value )) %>%
  rename(., housing_unit20 = value) 
  
# tract_tx_median_value <- tract_tx_median_value %>% select(., c("GEOID", estimate )) %>%
#   rename(., median_value20 = estimate) 
# 
# tract_tx_movers <- tract_tx_movers %>% select(., c("GEOID", estimate )) %>%
#   rename(., movers20 = estimate) 


merged_tract_tx <- tract_tx_pop_2020 %>%
    left_join(., tract_tx_hunit_2020, 
            by = c("GEOID" = "GEOID"))


#save the dataset as a CSV 
write.csv(merged_tract_tx, "~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/save_census_data_tract.csv") 


```


#Dataset 3 Pull population for Texas state then county


Get full list of variables in each year.
Full list of variables available at this URL : https://api.census.gov/data/2023/acs/acs5/profile/variables.html 
```{r}
acs13 <- load_variables(year = 2013,
                        dataset = "acs5", 
                        cache = TRUE)
```


```{r}
acs18 <- load_variables(year = 2018,
                        dataset = "acs5", 
                        cache = TRUE)
```

```{r}
acs23 <- load_variables(year = 2023,
                        dataset = "acs5", 
                        cache = TRUE)
```


```{r}
# Load variables from ACS 5-year profile
acs20_profile <- load_variables(year = 2020,
                        dataset = "acs5/profile", 
                        cache = TRUE)

acs10_profile <- load_variables(year = 2010,
                        dataset = "acs5/profile", 
                        cache = TRUE)


```

Pull population for Texas 
The population was 28,635,442 in 2020 and 24311891 in 2010.

```{r}
tx_pop <- get_acs(geography = "state",
                     variables = c("DP05_0001"),
                     state = "TX",
                     survey = "acs5",
                     year = 2020)

tx_pop10 <- get_acs(geography = "state",
                     variables = c("DP05_0001"),
                     state = "TX",
                     survey = "acs5",
                     year = 2010)

tx_pop00 <- get_decennial(geography = "state",
                 variables = c("P001001"),
                state = "TX",
                year = 2000)
```
2020 housing units DP04_0027 
The total number of housing units in 2023 was 11,112,975.
 In 2013, it was 10,070,703.
```{r}
tx_units <- get_acs(geography = "state",
                     variables = c("DP04_0027"),
                     state = "TX",
                     survey = "acs5",
                     year = 2020)


tx_units10 <- get_acs(geography = "state",
                     variables = c("DP04_0001"), #DP04_0027
                     state = "TX",
                     survey = "acs5",
                     year = 2010)

tx_unit00 <- get_decennial(geography = "state",
                 variables = c("H001001"),
                state = "TX",
                year = 2000)
```

```{r}
# median home value  2023 $260,400 
tx_value <- get_acs(geography = "state",
                     variables = c("B25077_001"),
                     state = "TX",
                     survey = "acs5",
                     year = 2023)

# median homevalue 2013 128900
tx_value13 <- get_acs(geography = "state",
                     variables = c("B25077_001"),
                     state = "TX",
                     survey = "acs5",
                     year = 2013)

value_change <- (tx_value$estimate - tx_value13$estimate )*100 / tx_value13$estimate
  
value_change
```


Pull ACS data from 2020
```{r}
# FETCH the data 
# DP05_0001 Total population 
# DP05_0091 number of housing units  # can confirm this is the same as total housing units (DP04)
# DP04_0001E total housing units in 2013 and 2023 but is depending on housing characteristics
# B25077_001 Median home value
# B07101_001E is the number of movers from another region... 

county_tx_pop <- get_acs(geography = "county",
                     variables = c("DP05_0001"),
                     state = "TX",
                     survey = "acs5",
                     year = 2020)

county_tx_hunit <- get_acs(geography = "county",
                 variables = c("DP04_0027"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2020)

# county_tx_median_value <- get_acs(geography = "county",
#                  variables = c("B25077_001"),
#                 state = "TX",
#               #  county = 201,
#                 survey = "acs5",
#                 year = 2023)

# DP04_0001 2023
county_tx_hunit_occ <- get_acs(geography = "county",
                 variables = c("DP04_0001"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2023) 

#all NA values... hmmm? 
# county_tx_movers <- get_acs(geography = "county",
#                  variables = c("B07101_001"),
#                 state = "TX",
#               #  county = 201,
#                 survey = "acs5",
#                 year = 2023)
#
```

Let's look at 10 years ago
```{r}
county_tx_pop_2010 <- get_acs(geography = "county",
                     variables = "DP05_0001",
                     state = "TX",
                     survey = "acs5",
                     year = 2010)

county_tx_hunit_2010 <- get_acs(geography = "county", 
                 variables = c("DP04_0001"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2010)

# county_tx_median_value_2013 <- get_acs(geography = "county",
#                  variables = c("B25077_001"),
#                 state = "TX",
#               #  county = 201,
#                 survey = "acs5",
#                 year = 2013)

```
Reach back to 2000
```{r}
county_tx_pop_2000 <- get_decennial(geography = "county",
                 variables = c("P001001"),
                state = "TX",
                year = 2000)

county_tx_hunit_2000 <- get_decennial(geography = "county", 
                 variables = c("H001001"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2000)
```



Remove the variables we don't need and merge to create the dataset 
a standard 254 counties across the 20 year time period...
```{r}
county_tx_pop <- county_tx_pop %>% select(., c("GEOID", "NAME", estimate )) %>%
  rename(., population20 = estimate) 

county_tx_hunit <- county_tx_hunit %>% select(., c("GEOID", estimate )) %>%
  rename(., housing_unit20 = estimate) 
  
# county_tx_median_value <- county_tx_median_value %>% select(., c("GEOID", estimate )) %>%
#   rename(., median_value23 = estimate) 

# county_tx_hunit_occ <- county_tx_hunit_occ %>% select(., c("GEOID", estimate )) %>%
#   rename(., unit_occ23 = estimate) 

county_tx_pop_2010 <- county_tx_pop_2010 %>% select(., c("GEOID", estimate )) %>%
  rename(., population10 = estimate) 

county_tx_hunit_2010 <- county_tx_hunit_2010 %>% select(., c("GEOID", estimate )) %>%
  rename(., housing_unit10 = estimate) 
# 
# county_tx_median_value_2013 <- county_tx_median_value_2013 %>% select(., c("GEOID", estimate)) %>%
#   rename(., median_value13 = estimate) 

county_tx_pop_2000 <- county_tx_pop_2000 %>% select(., c("GEOID", value )) %>%
  rename(., population00 = value) 

county_tx_hunit_2000 <- county_tx_hunit_2000 %>% select(., c("GEOID", value )) %>%
  rename(., housing_unit00 = value) 


merged_census_tx_county <- county_tx_pop %>%
    left_join(., county_tx_hunit, 
            by = c("GEOID" = "GEOID")) %>% 
    left_join(., county_tx_pop_2010, 
            by = c("GEOID" = "GEOID")) %>% 
    left_join(., county_tx_hunit_2010, 
            by = c("GEOID" = "GEOID")) %>%
    left_join(., county_tx_pop_2000, 
            by = c("GEOID" = "GEOID")) %>%
    left_join(., county_tx_hunit_2000, 
            by = c("GEOID" = "GEOID"))
  
```

```{r}
#save the dataset as a CSV 
write.csv(merged_census_tx_county, "~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/save_census_data_county_2000_10_20.csv") 

summary(merged_census_tx_county)
```


# Download counties from texas 

```{r}
# Get counties 
download.file("https://www2.census.gov/geo/tiger/TIGER2020PL/LAYER/COUNTY/2020/tl_2020_48_county20.zip", 
              destfile="Data/county_shape_48.zip")

county48 <-fs::dir_info(here::here("Data"  
                                 ))%>%  #$ means exact match
  dplyr::filter(str_detect(path, 
                           "tl_2020_48_county20.shp$"))%>%
  dplyr::select(path)%>%
  dplyr::pull()%>%
  
  sf::st_read() %>%
  st_transform(., 4326)
```


May 12  - Places dataset 
#sf1 has just the housing and occupancy variables
# Total units: H001001 
# owner occupied: H004002 
# total population... P001001

# I want the "dp" dataset from the census...
  # DP1_0001C is total population 
  # DP1_0147C is total housing units
  # DP1_0001C is total population for 2020
  
# for 2020 DHC 
 # P1_001N is the total population variable
 H1_001N total units variable
  
  
  
```{r}
summary_decennial <- load_variables(year = 2020,
                        dataset = "sf1", 
                        cache = TRUE)


ddhcb_summary_decennial <- load_variables(year = 2020,
                        dataset = "dhc", 
                        cache = TRUE)
```

```{r}
#the places change over time... that's okay! just need to figure out the subset that I am tracking...
#pull the population counts 
places_tx_pop_2000 <- get_decennial(geography = "place",
                 variables = c("P001001"),
                state = "TX",
              #  county = 201,
               # survey = "acs5",
                year = 2000)

places_tx_pop_2010 <- get_decennial(geography = "place",
                 variables = c("P001001"),
                state = "TX",
              #  county = 201,
               # survey = "acs5",
                year = 2010)

places_tx_pop_2020 <- get_decennial(geography = "place",
                 variables = c("P1_001N"),
                state = "TX",
                year = 2020)


#pull the housing unit counts 
places_tx_hunit_2000 <- get_decennial(geography = "place",
                 variables = c("H001001"),
                state = "TX",
                year = 2000)

places_tx_hunit_2010 <- get_decennial(geography = "place",
                 variables = c("H001001"),
                state = "TX",
                year = 2010)

places_tx_hunit_2020 <- get_decennial(geography = "place",
                 variables = c("H1_001N"),
                state = "TX",
                year = 2020)



#rename and reduce the population
places_tx_pop_2000 <- places_tx_pop_2000 %>% select(., c("GEOID", "NAME", value )) %>%
  rename(., pop00 = value) 

places_tx_pop_2010 <- places_tx_pop_2010 %>% select(., c("GEOID", "NAME", value )) %>%
  rename(., pop10 = value)

places_tx_pop_2020 <- places_tx_pop_2020 %>% select(., c("GEOID", "NAME", value )) %>%
  rename(., pop20 = value)

#rename and reduce the housing units 
places_tx_hunit_2000 <- places_tx_hunit_2000 %>% select(., c("GEOID", value )) %>%
  rename(., hunit00 = value) 

places_tx_hunit_2010 <- places_tx_hunit_2010 %>% select(., c("GEOID", value )) %>%
  rename(., hunit10 = value) 

places_tx_hunit_2020 <- places_tx_hunit_2020 %>% select(., c("GEOID", value )) %>%
  rename(., hunit20 = value) 


#merge and take left join, so only the places that existed at time of 2000.... 
merged_census_places_tx <- places_tx_pop_2000 %>%
    left_join(., places_tx_hunit_2000, 
          by = c("GEOID" = "GEOID")) 

merged_census_places_tx_10 <- places_tx_pop_2010 %>%
    left_join(., places_tx_hunit_2010, 
          by = c("GEOID" = "GEOID")) 

merged_census_places_tx_20 <- places_tx_pop_2020 %>%
    left_join(., places_tx_hunit_2020, 
          by = c("GEOID" = "GEOID")) 
  
```

```{r}
#only has the 1510 places that 2000 existed...
all_years_census_places_tx <- merged_census_places_tx %>%
    left_join(., merged_census_places_tx_10, 
          by = c("GEOID" = "GEOID")) %>%
    left_join(., merged_census_places_tx_20, 
          by = c("GEOID" = "GEOID")) 
```

#save and write the data...
```{r}
#save the dataset as a CSV 
write.csv(all_years_census_places_tx, "~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/save_census_data_places.csv") 
```

#Dataset 5 or 6 - (May 15) New datset to get the census tract population and housing units difference from 2010 to 2019 

```{r}
summary_acs <- load_variables(year = 2010,
                        dataset = "acs5", 
                        cache = TRUE)

#B01003_001 for population 
tract_tx_pop_2010 <- get_acs(geography = "tract",
                 variables = c("B01003_001"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2010)
tract_tx_pop_2019 <- get_acs(geography = "tract",
                 variables = c("B01003_001"),
                state = "TX",
                survey = "acs5",
                year = 2019)

# B25001_001 for housing units 
tract_tx_hunit_2010 <- get_acs(geography = "tract",
                 variables = c("B25001_001"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2010)

tract_tx_hunit_2019 <- get_acs(geography = "tract",
                 variables = c("B25001_001"),
                state = "TX",
                survey = "acs5",
                year = 2019)

```
```{r}
tract_tx_pop_2010 <- tract_tx_pop_2010 %>% select(., c("GEOID", "NAME", estimate )) %>%
  rename(., pop10 = estimate)

tract_tx_pop_2019 <- tract_tx_pop_2019 %>% select(., c("GEOID", estimate )) %>%
  rename(., pop19 = estimate)

tract_tx_hunit_2010 <- tract_tx_hunit_2010 %>% select(., c("GEOID",  estimate )) %>%
  rename(., hunit10 = estimate)

tract_tx_hunit_2019 <- tract_tx_hunit_2019 %>% select(., c("GEOID", estimate )) %>%
  rename(., hunit19 = estimate)


all_tract_2010_2019 <- tract_tx_pop_2010 %>%
    left_join(., tract_tx_pop_2019, 
          by = c("GEOID" = "GEOID")) %>%
    left_join(., tract_tx_hunit_2010, 
          by = c("GEOID" = "GEOID")) %>%
    left_join(., tract_tx_hunit_2019, 
          by = c("GEOID" = "GEOID")) 
```

```{r}
#save the dataset as a CSV 
write.csv(all_tract_2010_2019, "~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/save_census_tracts_2010_2019.csv") 
```

