---
title: "Places_deepdive"
output: html_document
date: "2025-02-24"
---

Map the specific metro areas in Texas

```{r Load Libraries}
library(tidycensus)
library(tidyverse)
library(choroplethr)
library(choroplethrMaps)
library(ggplot2)
library(dplyr)
library(tmap)
library(sf)
```

```{r Additional libraries}
#specific for data pull
library(jsonlite)

library(janitor)
```

Steps:
-- Pull Data:
    --places boundaries, redlining
-- select one place, intersect with the tracts layer 
-- Project historic redlining

Pull the 'place' boundary according to the US Census 
```{r}
#Definine which city I am looking into 
city <- "Austin" #"Houston" # "Dallas"

places <- st_read("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/tl_2024_48_place/tl_2024_48_place.shp") %>% st_transform(., 4326)

```

# Context questions to answer: 
  - Why Houston? Did Dallas too
  - Why do they have a really weird outline? why did I extend the intersect boundary by 100 miles? 

Project the places boundary 
```{r pressure, echo=FALSE}
city_boundary  <- places %>% filter(., NAME == city)
qtm(city_boundary)
```

Pull historic redlining

```{r}

geojson_url <- paste("https://services.arcgis.com/ak2bo87wLfUpMrt1/ArcGIS/rest/services/MappingInequalityRedliningAreas_231211/FeatureServer/0/query?where=city+%3D+%27", toupper(city), "%27&objectIds=&geometry=&geometryType=esriGeometryEnvelope&inSR=&spatialRel=esriSpatialRelIntersects&resultType=none&distance=0.0&units=esriSRUnit_Meter&relationParam=&returnGeodetic=false&outFields=*&returnGeometry=true&returnCentroid=false&returnEnvelope=false&featureEncoding=esriDefault&multipatchOption=xyFootprint&maxAllowableOffset=&geometryPrecision=&outSR=&defaultSR=&datumTransformation=&applyVCSProjection=false&returnIdsOnly=false&returnUniqueIdsOnly=false&returnCountOnly=false&returnExtentOnly=false&returnQueryGeometry=false&returnDistinctValues=false&cacheHint=false&collation=&orderByFields=&groupByFieldsForStatistics=&outStatistics=&having=&resultOffset=&resultRecordCount=&returnZ=false&returnM=false&returnTrueCurves=false&returnExceededLimitFeatures=true&quantizationParameters=&sqlFormat=none&f=pgeojson&token=", sep = "")

redlining_boundary <- st_read(geojson_url)  %>% st_transform(., 4326)

```
Project the redlining boundary 
```{r}
#figure out some casting / shaping of the json object... 
qtm(redlining_boundary$geometry)
```

#Plot the boundaries and the redlining
```{r}
redlining_boundary <- redlining_boundary[st_intersects(redlining_boundary, city_boundary) %>% lengths > 0,]

ggplot() + 
  geom_sf(data = city_boundary, aes(geometry = geometry)) +
   geom_sf(data = redlining_boundary, aes(fill = fill, geometry = geometry))  +
  labs(
    title = paste(city, "Texas Boundaries Overlayed with Historic Redlining"),
    caption = "Data: Redlining Areas, University of Richmond | US Census"
  ) + # +  theme(legend.title = c("Redlining Areas", ""))
theme_classic()

# + scale_fill_manual(labels = redlining_boundary$category)

#values = c('darkgreen', 'red', 'yellow'),

```

Load the census tracts 
```{r}
texas_tracts <- st_read("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/tract_shape_48/tl_2020_48_tract20.shp") %>%
  st_transform(., 4326)
```

Create a buffer around census tracts to consider commuting zone 
```{r}
#add a 100km buffer
boundary_with_buffer <- st_buffer(city_boundary$geometry, 10000)

#TODO: Want the entire tract area of the ones that intersect... 
tracts_intersect_buffer <- texas_tracts[st_intersects(texas_tracts, boundary_with_buffer) %>% lengths > 0,]
 #<- st_intersection(texas_tracts, boundary_with_buffer)
```

#lets map them overlaid to see where the city covers the texas census tracts 

```{r}

ggplot() +  geom_sf(data = texas_tracts, aes(geometry = geometry)) +
 #  geom_sf(data = boundary_with_buffer, aes(geometry = geometry)) +
   geom_sf(data = tracts_intersect_buffer, fill = "red", aes(geometry = geometry)) 

```

Proceed with just mapping the tracts and the overlay of the 'place' boundary

Plot the buffer with the boundary too 
```{r}
ggplot()  +  geom_sf(data = tracts_intersect_buffer, color = "grey", fill = NA, aes(geometry = geometry)) + 
    geom_sf(data = city_boundary, fill = "red", color= "red", alpha = 0.1, aes(geometry = geometry)) + 
  theme_void()

```


#Read in preprocessed data
```{r}
#save the dataset as a CSV 
merged_tract_tx <- read_csv("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/save_census_data_tract.csv") %>%
  select(., c("GEOID", "population23", "housing_unit23", "median_value23"))

summary(merged_tract_tx)
```

```{r}
tract_tx_total_hunit <- read_csv("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/housing_tract_census_tx.csv") %>% 
  select(., c("GEOID", "units_built_2010_19", "units_built_2020", 
              "units_built_since_2010"))

tract_tx_total_hunit$GEOID <- as.character(tract_tx_total_hunit$GEOID)

summary(tract_tx_total_hunit)

merged_tract_tx <- merged_tract_tx %>%
  left_join(., tract_tx_total_hunit, by = c("GEOID" = "GEOID"))
```


```{r}
summary(merged_tract_tx)
```

```{r}
pairs(merged_tract_tx %>% select(2 :7))

```

```{r}
hist(merged_tract_tx$population23)

hist(merged_tract_tx$units_built_since_2010)

```


#Merge the census data with the entire state shapefile of census tracts....

```{r}
merged_tract_tx$GEOID <- as.character(merged_tract_tx$GEOID)

merged_tract_tx_geo <-  merged_tract_tx %>% 
  left_join(., texas_tracts, by = c("GEOID" = "GEOID20"))
```

##todo:
 - calculate the housing density at the tract level, population density, housing growth density
 
```{r}
merged_tract_tx_geo <- merged_tract_tx_geo %>% 
  mutate(pop_density = population23/ALAND20 ) %>%
  mutate(housing_density = housing_unit23/ALAND20) %>%
  mutate(unit_growth_density_2010 = units_built_since_2010/ALAND20)
  
```
 

# lets look at housing production by 
```{r}
ggplot() + geom_sf(data = merged_tract_tx_geo, aes(fill = pop_density, geometry = geometry),  lwd = 0)  +
  theme_void() + 
    scale_fill_distiller(palette="YlGn" , direction = 1)

```

```{r}
ggplot() + geom_sf(data = merged_tract_tx_geo, aes(fill = housing_density, geometry = geometry),  lwd = 0)  +
  theme_void() + 
    scale_fill_distiller(palette="YlGn" , direction = 1)
```
```{r}
ggplot() + geom_sf(data = merged_tract_tx_geo, aes(fill = unit_growth_density_2010, geometry = geometry),  lwd = 0)  +
  theme_void() + 
    scale_fill_distiller(palette="YlGn" , direction = 1)
```

## should I do some autocorrelation on the counties? Yes 
## map state wide analysis... 



```{r}
#merge the census data with the boundaries 

tracts_intersect_buffer <- tracts_intersect_buffer %>% 
  left_join(., merged_tract_tx_geo, by = c("GEOID20" = "GEOID"))

#tracts_intersect_buffer <- tracts_intersect_buffer %>%
#  mutate(unit_density = (housing_unit23.x / ALAND20) )

ggplot()  +  geom_sf(data = tracts_intersect_buffer, aes(fill = housing_density.x, geometry = geometry.x)) +
#+ geom_sf(data = city_boundary, fill = "red", color= "red", alpha = 0.1, aes(geometry = geometry)) + 
     geom_sf(data = city_boundary, fill = NA, color= "red", alpha = 0.1, aes(geometry = geometry)) + 
  theme_classic() +
  scale_fill_distiller(palette="Greens" , direction = 1) + 
  labs(
    title = paste(city, "Texas Census Tracts with unit density and overlay of city outline"),
    subtitle = "ACS 2023",
    caption = "Data: US Census | ACS"
  )
```

```{r}
ggplot()  +  geom_sf(data = tracts_intersect_buffer, aes(fill = unit_growth_density_2010.x, geometry = geometry.x)) +
#+ geom_sf(data = city_boundary, fill = "red", color= "red", alpha = 0.1, aes(geometry = geometry)) + 
     geom_sf(data = city_boundary, fill = NA, color= "red", alpha = 0.1, aes(geometry = geometry)) + 
  theme_classic() +
  scale_fill_distiller(palette="Greens" , direction = 1) + 
  labs(
    title = paste(city, "Texas Census Tracts with "),
    subtitle = "Unit density growth and overlay of city outline",
    caption = "Data: US Census | ACS 5 year 2023"
  )
```



Introduce the Climate risk data ... 
```{r}
NRI_tract <- read_csv("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/National_Risk_Index_Census_Tracts_-2688098782675405028.csv")

census_NRI <- st_read("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/National_Risk_Index_Census_Tracts_-2688098782675405028.csv")

#cut it down to texas
NRI_tract <- NRI_tract %>% clean_names()

#includes data for 6883 tracts out of 6896 tracts
NRI_tx_tract <- NRI_tract %>% filter(., state_fips_code == 48)
 
NRI_tx_tract_small <-  NRI_tx_tract %>% select(., c("county_name", "census_tract", "census_tract_fips_code", "population_2020", "building_value", "area_sq_mi", ,"national_risk_index_value_composite", "national_risk_index_score_composite"))


#try plotting the tracts visually... 

tracts_intersect_buffer <- tracts_intersect_buffer %>%
  left_join(., NRI_tx_tract_small, by = c("GEOID20" = "census_tract_fips_code"))
```

```{r}
ggplot()  +  geom_sf(data = tracts_intersect_buffer, aes(fill = national_risk_index_value_composite, geometry = geometry.x)) +
#+ geom_sf(data = city_boundary, fill = "red", color= "red", alpha = 0.1, aes(geometry = geometry)) +
  theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1) + 
  labs(
    title = paste(city, "Texas Census Tracts with NRI"),
    subtitle = "ACS 2023",
    caption = "Data: US Census | NRI FEMA"
  )
```

```{r}
## lets plot unit density versus NRI 

ggplot() + geom_point(data = tracts_intersect_buffer, aes(y = unit_growth_density_2010.x, x= national_risk_index_value_composite), alpha = 1 / 3) +
    geom_smooth(data = tracts_intersect_buffer, aes(y=unit_growth_density_2010.x, x = national_risk_index_value_composite), method = "lm", se = FALSE)



```

```{r}
#how much housing stock is already in the high risk area? 
#is the housing growth disproportionately happening there? 

#could we look at percentage outside versus inside the incorporated area? 


#built since 2010 239,156 , 827,682 total in the area 
sum(tracts_intersect_buffer$units_built_since_2010.x)
sum(tracts_intersect_buffer$housing_unit23.x)

tracts_intersect_buffer$NRI_quartile <- ntile(tracts_intersect_buffer$national_risk_index_value_composite, 5)


summary_austin <- tracts_intersect_buffer %>% group_by(NRI_quartile) %>% 
  summarise(num = n(), totalunit = sum(housing_unit23.x), total_since_2010 = sum(units_built_since_2010.x), total_land = sum(ALAND20))

summary_austin <- summary_austin %>% 
  mutate(total_unit_density = totalunit / total_land) %>%
  mutate(growth_density = total_since_2010 / total_land)

summary_austin
```

