---
title: "Census_Data_pull"
output: html_document
date: "2025 Spring"
---
Script to pull Census Data and map
Map the state of texas and also analyse the counties... 

```{r}
# Save census API key
API_Key = '83a719c9a57b7c2fedd093f9a0108fcf8bf552eb'

```

Pull data from the census website for population, boundaries, etc .

Try the tidy census package from the census API development instructions
https://walker-data.com/tidycensus/articles/basic-usage.html
#API for census data https://www.census.gov/data/developers/data-sets.html

#https://edits.nationalmap.gov/apps/gaz-domestic/public/search/names 
# texas is code 48 places look up through USGS 
 
```{r}
library(tidycensus)
library(tidyverse)
library(choroplethr)
library(choroplethrMaps)
library(ggplot2)
library(dplyr)
library(tmap)
library(sf)
library(terra)
```


```{r}
#pull in the dataset if not generated...

merged_census_tx <- read.csv("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/save_census_data_county_2000_10_20.csv")
```

TODO: #could also calculate a population or housing unit density.. 

Perform calculations to find the change in variables over 10 years
```{r}
#add new column from 2013 population estimate and also calcualte percentage change over the 10 years
#rename column 

merged_census_tx <- merged_census_tx %>%
  mutate(popchange10 = 
  (merged_census_tx$population20 - merged_census_tx$population10)*100 / merged_census_tx$population10)%>%
  mutate(unitchange10 = (housing_unit20 - housing_unit10)*100 / housing_unit10)

 # merged_census_tx <- merged_census_tx %>% mutate(area = st_area(merged_census_tx$geometry))

#  merged_census_tx <- merged_census_tx %>% mutate(pop_density = #population23/st_area(merged_census_tx$geometry))
```



```{r}
#create a matrix plot of scatter plots to briefly understand the relationships between variables
#pairs(merged_census_tx %>% select(4:13))
```

#visualize the population by county, shows there are very few with a big population... 
# {r}
# merged_census_tx %>%
#   ggplot(aes(y = population23, x = reorder(NAME, population23))) +
#   geom_point()

Plot the growth over 10 years 
```{r}
# merged_census_tx %>%
#   ggplot(aes(y = popchange10, x = reorder(NAME, popchange10))) +
#   geom_point()
```
Find five fastest growing counties since 2013

```{r}
#get the top values of tp_tx_county$tenyeargrowth 
#TODO: put the 2013 estimate on the dataframe... 
summary(merged_census_tx$popchange10)
#which values were these? 
tail(merged_census_tx[order(merged_census_tx$popchange10), ], 7)

```


Load in geography 

```{r}

#if already downloaded texas counties... 
county48 <- st_read("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/tl_2020_48_county20.shp")

qtm(county48)
```
let's use ggplot2 to visualise... 


```{r}
# create a merged dataset with the geometry... 
merged_census_tx$GEOID <- as.character(merged_census_tx$GEOID)

texas_geo_subset <- county48 %>% select("GEOID20", "geometry")
merged_texas <- merged_census_tx %>%
  left_join(texas_geo_subset, by = c("GEOID" = "GEOID20"))

```

#Plot the percent change of population across Counties
```{r}
ggplot(merged_texas) +
  geom_sf(aes(fill = popchange10, geometry = geometry)) +
    labs(
    title = "Population Change Across Texas Counties",
    subtitle = "Percent Change from 2010 to 2020",
    caption = "Data: American Community Survey",
     fill = "Percent Population Change"
  ) +
  theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1)
```

```{r}
ggplot(merged_texas) +
  geom_sf(aes(fill = unitchange10, geometry = geometry)) +
    labs(
    title = "Housing Unit Change Across Texas Counties",
    subtitle = "Percent Change from 2010 to 2020",
    caption = "Data: American Community Survey", 
     fill = "Percent Unit Change"
  ) +
  theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1) 
```

```{r}
#Questions: where did population grow and housing units did not?
#where did population grow and how did that compare to risk?

#which are the counties with the highest population change and highest risk score? 
highest_pop_change <- tail(merged_texas[order(merged_texas$popchange10), ], 7)
highest_unit_change <- tail(merged_texas[order(merged_texas$unitchange10), ], 7)
#highest_value_change <- tail(merged_texas[order(merged_texas$valuechange10, na.last = FALSE), ], 5)

top_quintile_pop_change <- tail(merged_texas[order(merged_texas$popchange10), ], round(count(merged_texas) / 5)[[1]])

summary(merged_texas$popchange10)

highest_unit_change
```

```{r}
#map again 
ggplot() +
  geom_sf(data= merged_texas, aes(fill = popchange10, geometry = geometry)) +
    geom_sf(data= highest_pop_change, aes(color = NAME, fill = popchange10, geometry = geometry)) +
    labs(
    title = "Population Change Across Texas Counties",
    subtitle = "Percent Change of 2010 to 2020, top seven outlined in red",
    caption = "Data: American Community Survey", 
    fill=  "Percent Population Change", 
    color = "Counties") +
  theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1)
```

```{r}
#TODO next: 
# map the climate indexes and the real estate prices
# try mapping only an area but down to the tract level (ie. Houston and Harriscounty or sth...)
  # then try doing some correlation? 

#Write up some introduction? 
```

Load in the National Risk INDEX (NRI)
```{r}
NRI_dictionary <- read_csv("~/Documents/_UCL_grad school/UCL/Dis/NRI_Table_Counties_Texas/NRIDataDictionary.csv")
```

```{r}
NRI_info <- read_csv("~/Documents/_UCL_grad school/UCL/Dis/NRI_Table_Counties_Texas/NRI_HazardInfo.csv")

NRI_tx <- read_csv("~/Documents/_UCL_grad school/UCL/Dis/NRI_Table_Counties_Texas/NRI_Table_Counties_Texas.csv")
```
#TODO: 
  Some additional questions: 
  - what is the distribution of the risk_score?  --> assuming its normalized to fit a 0-100 scale? 
  - how does population compare to the population from ACS? Very similar 
  - distribution of buildvalue? 
```{r}

NRI_tx_small <- NRI_tx %>% select("COUNTY", "COUNTYTYPE", "COUNTYFIPS", "STCOFIPS", "POPULATION", "BUILDVALUE",
                                  "AGRIVALUE", "RISK_VALUE", "RISK_SCORE", "EAL_VALB" ,"ALR_VALB") %>%
              mutate( STCOFIPS = as.character(STCOFIPS))
#could add additional variables about expected loss
```


```{r}
summary(NRI_tx_small)
print(mean(NRI_tx_small$RISK_VALUE))
```

```{r}
ggplot(NRI_tx_small, aes(x=RISK_VALUE)) + geom_histogram(binwidth=10000, colour="black", fill="white")


ggplot(NRI_tx_small, aes(x=RISK_SCORE)) + geom_histogram(binwidth=10, colour="black", fill="white")



```
Whats the distribution for the building loss rate? 
Percentage of total building value 
```{r}
ggplot(NRI_tx_small, aes(x=ALR_VALB)) + geom_histogram(binwidth=.00010, colour="black", fill="white")

```

```{r}
#create dataset with the subset of NRI variables
merged_texas <- merged_texas %>%
            left_join(NRI_tx_small, by = c("GEOID"= "STCOFIPS"))
       #      %>% select(., c("iso3", "country", "hdicode", "region",  "gii_2019", "gii_2020")) 
```

###create dataset with all the NRI variables so I can map them... 
```{r}
merged_texas_all_NRI <- merged_texas %>%
            left_join(NRI_tx %>% mutate( STCOFIPS = as.character(STCOFIPS)), by = c("GEOID"= "STCOFIPS"))
       #      %>% select(., c("iso3", "country", "hdicode", "region",  "gii_2019", "gii_2020")) 
```


```{r}
world_cites <- st_read("https://services.arcgis.com/P3ePLMYs2RVChkJx/arcgis/rest/services/World_Cities/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson") %>%
  st_transform(., 4326)

texas_cities <- world_cites %>% filter(., FIPS_CNTRY == "US" & ADMIN_NAME == "Texas")

```

```{r}
q <- ggplot()+  geom_sf(data = texas_cities, aes(fill = "black", geometry = geometry )) + 
  geom_sf_text(data = texas_cities, aes(label = CITY_NAME, geometry = geometry), position = position_nudge(x = 0.6, y = 0.2)) +
 # geom_text(data = texas_cities, aes(label = CITY_NAME), size = 5) + 
  theme_void()
```


Regressions 
```{r}


#positive correlation between both the risk and the units... 
lm_fit <- lm(unitchange10 ~ RISK_SCORE, data=merged_texas)
summary(lm_fit)

lm_fit <- lm(popchange10 ~ RISK_SCORE, data=merged_texas)
summary(lm_fit)

lm_fit <- lm(valuechange10 ~ RISK_SCORE, data=merged_texas)
summary(lm_fit)

lm_fit <- lm(valuechange10 ~ popchange10, data=merged_texas)
summary(lm_fit)

ggplot() +  geom_point(data = merged_texas, aes(x =RISK_SCORE, y = unitchange10), alpha = 1 / 3) +
    geom_point(data = highest_unit_change, aes(x =  RISK_SCORE,  y= unitchange10), color = "Red" ) + 
  geom_smooth(data = merged_texas, aes(x =RISK_SCORE, y = unitchange10), method = "lm", se = FALSE)

#+ xlab("NRI Risk Score") + ylab("10 year Housing Unit change", )

ggplot() + geom_point(data = merged_texas, aes(x = RISK_SCORE, y= popchange10), alpha = 1 / 3) +
  #xlab("NRI Risk Score") + ylab("10 year population change") +  geom_point(alpha = 1 / 3) +
    geom_point(data = highest_pop_change, aes(x =  RISK_SCORE,  y= popchange10), color = "Red" ) + 
    geom_smooth(data = merged_texas, aes(x=RISK_SCORE, y = popchange10), method = "lm", se = FALSE) #+
  #   geom_label(data = highest_pop_change, aes(x =  RISK_SCORE,  y= popchange10), label = highest_pop_change$COUNTY, nudge_x=0.45, nudge_y=0.1, check_overlap=T)


ggplot() + geom_point(data = merged_texas, aes(x = RISK_SCORE, y= valuechange10), alpha = 1/3) + 
  #xlab("NRI Risk Score") + ylab("10 year median house value change") +  geom_point(alpha = 1 / 3)  +
    geom_point(data = highest_value_change, aes(x =  RISK_SCORE,  y= valuechange10), color = "Red" ) +
    geom_smooth(data = merged_texas, aes(x = RISK_SCORE, y= valuechange10), method = "lm", se = FALSE) #+
 #  geom_label(data = highest_value_change, aes(x =  RISK_SCORE,  y= valuechange10), label = highest_value_change$COUNTY, nudge_x=1, nudge_y=1, check_overlap=T)


```


```{r}
high_scoring <- merged_texas[(merged_texas$unitchange10 >40 & merged_texas$RISK_SCORE > 75 ),]


```

#TODO Wednesday: 
#something about adding the high scoring ones in highlight or with labels...? 
#then also add them with maps - highlight those counties
- label the known cities?


#identify coastal?
#also map the flooding? 

## Todo: add labels with the names of the cities... and then can we get the values? 
```{r}

ggplot() +
  geom_sf(data = merged_texas, aes(fill = RISK_SCORE, geometry = geometry)) +
  #geom_sf(data = texas_cities, aes(fill = "black", geometry = geometry)) 
#    geom_sf_text(data = texas_cities, aes(label = CITY_NAME, geometry = geometry), position = position_nudge(x = 0.6, y = 0.2)) +
  #    geom_sf(data= top_quintile_pop_change, aes(color = "red",fill = RISK_SCORE, geometry = geometry)) +
labs(
    title = "Risk Score Across Texas Counties",
    subtitle = "National Risk Index March 2023",
    caption = "Data: FEMA National Risk Index | US Census", 
    fill = "Risk Score"
  ) + 
  theme(
    plot.title = element_text(size = 24, hjust = 0.5)
  ) +
    theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1) 

```

#create additional maps of the hazard scores.. .using merged_texas_all_NRI
```{r}
#EAL_VALT - total expected annual loss
#VARIABLE = CLFD_EALB

ggplot() +
  geom_sf(data = merged_texas_all_NRI, aes(fill = ALR_VALA, geometry = geometry)) +
  #geom_sf(data = texas_cities, aes(fill = "black", geometry = geometry)) 
#    geom_sf_text(data = texas_cities, aes(label = CITY_NAME, geometry = geometry), position = position_nudge(x = 0.6, y = 0.2)) +
    #  geom_sf(data= top_quintile_pop_change, aes(color = "red",fill = RISK_SCORE, geometry = geometry)) +
labs(
    title = "Coastal Flooding Risk Across Texas Counties",
    subtitle = "National Risk Index March 2023, Counties in top Quintile of population growth in red",
    caption = "Data: FEMA National Risk Index | ESRI"
  ) + 
  theme(
    plot.title = element_text(size = 24, hjust = 0.5)
  ) +
    theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1) 

```


Autocorrelation NRI for the whole state of texas and the counties... 
```{r}
library(spdep)

coordsW <- merged_texas$geometry%>%
  st_centroid()%>%
  st_geometry()
  
plot(coordsW,axes=TRUE)


#create a neighbours list
LWard_nb <- merged_texas$geometry %>%
  poly2nb(., queen=T)

summary(LWard_nb)


#plot them
plot(LWard_nb, st_geometry(coordsW), col="red")
#add a map underneath
plot(merged_texas$geometry, add=T)

#create a spatial weights matrix from these weights
Lward.lw <- LWard_nb %>%
  nb2mat(., style="B")

sum(Lward.lw)

sum(Lward.lw[1,])

Lward.lw <- LWard_nb %>%
  nb2listw(., style="C")

I_LWard_Global_Density <- merged_texas %>%
  pull(RISK_SCORE) %>%
  as.vector()%>%
  moran.test(., Lward.lw)

I_LWard_Global_Density


G_LWard_Global_Density <- 
  merged_texas %>%
  pull(RISK_SCORE) %>%
  as.vector()%>%
  globalG.test(., Lward.lw)

G_LWard_Global_Density
```

```{r}
Gi_LWard_Local_Density <- merged_texas %>%
  pull(RISK_SCORE) %>%
  as.vector()%>%
  localG(., Lward.lw)

head(Gi_LWard_Local_Density)

merged_texas <- merged_texas %>%
  mutate(density_G = as.numeric(Gi_LWard_Local_Density))
```

#Map the GI autocorrelation

```{r}
library(RColorBrewer)

GIColours<- rev(brewer.pal(8, "RdBu"))

breaks1<-c(-1000,-2.58,-1.96,-1.65,1.65,1.96,2.58,1000)


#now plot on an interactive map
# tm_shape(merged_texas$geometry) +
#     tm_polygons(col = merged_texas$density_G,
#         style="fixed",
#         breaks=breaks1,
#         palette=GIColours,
#         midpoint=NA,
#         title="Gi*, NRI texas")


ggplot() +
  geom_sf(data = merged_texas, aes(fill = density_G, geometry = geometry)) +
  labs(
    title = "Risk Score Autocorrelation Across Texas Counties",
    subtitle = "Climate Event Risk Index March 2023",
    caption = "Data: FEMA National Risk Index | US Census", 
    fill = "Density Metric"
  ) + 
  theme(
    plot.title = element_text(size = 24, hjust = 0.5)
  ) +
    theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1) 
```


<!-- Let's map Harris county (Aka Houston ) -->
<!-- ```{r} -->
<!-- #Get the tracts first  -->
<!-- download.file("https://www2.census.gov/geo/tiger/TIGER2020PL/LAYER/TRACT/2020/tl_2020_48_tract20.zip",  -->
<!--               destfile="Data/tract_shape_48.zip") -->

<!-- # texas_tracts <-fs::dir_info(here::here("Data"   -->
<!-- #                                  ))%>%  #$ means exact match -->
<!-- #   dplyr::filter(str_detect(path,  -->
<!-- #                            "tract_shape_48/tl_2020_48_tract20.shp$"))%>% -->
<!-- # ##  dplyr::select(path)%>% -->
<!-- #   dplyr::pull()%>% -->
<!-- #   sf::st_read() -->

<!-- texas_tracts <- st_read("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/tract_shape_48/tl_2020_48_tract20.shp") %>% -->
<!--   st_transform(., 4326) -->

<!-- qtm(texas_tracts) -->


<!-- ``` -->


<!-- ```{r} -->
<!-- tp_hctx -->
<!-- ``` -->


```{r}
print(mean(merged_texas$popchange10)) #6.551286
print(sd(merged_texas$popchange10)) # 17.05852
merged_texas$popchange10z <- (merged_texas$popchange10-mean(merged_texas$popchange10))/sd(merged_texas$popchange10)

```

## Try to map the z score or standard deviation... 
```{r}
ggplot() +
  geom_sf(data = merged_texas, aes(fill = popchange10z, geometry = geometry)) +
#  geom_sf(data = texas_cities, aes(fill = "black", geometry = geometry )) +
  theme_classic() +
  labs(
    title = "Population Change Across Texas Counties",
    subtitle = "Standard deviation of 2010 to 2020",
    caption = "Data: American Community Survey",
    fill = "Standard Deviation"
  ) + 
   theme(
    plot.title = element_text(size = 24, hjust = 0.5)
  ) +
    theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1) 
```

#Autocorrelation of population change 2010 to 2020
```{r}
#Can reuse the neighbors list from above because geometry isnt changing

#calculate density based on popchange10
I_LWard_Global_Density_pop <- merged_texas %>%
  pull(popchange10) %>%
  as.vector()%>%
  moran.test(., Lward.lw)

I_LWard_Global_Density_pop


G_LWard_Global_Density_pop <- 
  merged_texas %>%
  pull(popchange10) %>%
  as.vector()%>%
  globalG.test(., Lward.lw)

G_LWard_Global_Density_pop
```

```{r}
Gi_LWard_Local_Density_pop <- merged_texas %>%
  pull(popchange10) %>%
  as.vector()%>%
  localG(., Lward.lw)

head(Gi_LWard_Local_Density_pop)

merged_texas <- merged_texas %>%
  mutate(density_G_pop = as.numeric(Gi_LWard_Local_Density_pop))
```

#Map the GI autocorrelation for population change

```{r}
GIColours<- rev(brewer.pal(8, "RdBu"))

breaks1<-c(-1000,-2.58,-1.96,-1.65,1.65,1.96,2.58,1000)


ggplot() +
  geom_sf(data = merged_texas, aes(fill = density_G_pop, geometry = geometry)) +
  labs(
    title = "Population Change Autocorrelation Across Texas Counties",
    subtitle = "Population Change 2010-2020",
    caption = "Data: US Census", 
    fill = "Density Metric"
  ) + 
  theme(
    plot.title = element_text(size = 24, hjust = 0.5)
  ) +
    theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1) 
```

#Autocorrelation of HOUSING change 2010 to 2020
```{r}
#Can reuse the neighbors list from above because geometry isnt changing

#calculate density based on unitchange10
I_LWard_Global_Density_unit <- merged_texas %>%
  pull(unitchange10) %>%
  as.vector()%>%
  moran.test(., Lward.lw)

I_LWard_Global_Density_unit


G_LWard_Global_Density_unit <- 
  merged_texas %>%
  pull(unitchange10) %>%
  as.vector()%>%
  globalG.test(., Lward.lw)

G_LWard_Global_Density_unit
```

```{r}
Gi_LWard_Local_Density_unit <- merged_texas %>%
  pull(unitchange10) %>%
  as.vector()%>%
  localG(., Lward.lw)

head(Gi_LWard_Local_Density_unit)

merged_texas <- merged_texas %>%
  mutate(density_G_unit = as.numeric(Gi_LWard_Local_Density_unit))
```



#Map the GI autocorrelation for population change

```{r}
GIColours<- rev(brewer.pal(8, "RdBu"))

breaks1<-c(-1000,-2.58,-1.96,-1.65,1.65,1.96,2.58,1000)


ggplot() +
  geom_sf(data = merged_texas, aes(fill = density_G_unit, geometry = geometry)) +
  labs(
    title = "Housing Unit Change Autocorrelation Across Texas Counties",
    subtitle = "Housing Unit Change 2010-2020",
    caption = "Data: US Census", 
    fill = "Density Metric"
  ) + 
  theme(
    plot.title = element_text(size = 24, hjust = 0.5)
  ) +
    theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1) 
```


#Put together some line graphs... 
```{r}
highest_pop_change_graph <- highest_pop_change %>%
  select(., c("NAME", "population00", "population10", "population20"))
#rename some of the columns
highest_pop_change_graph <-  rename(highest_pop_change_graph, "County" = "NAME", 
        "2000" = "population00", "2010" = "population10", "2020" = "population20")


highest_unit_change_graph <- highest_unit_change %>%
  select(., c("NAME", "housing_unit00", "housing_unit10", "housing_unit20"))

highest_unit_change_graph <-  rename(highest_unit_change_graph, "County" = "NAME",
        "2000" = "housing_unit00", "2010" = "housing_unit10", "2020" = "housing_unit20")
 

highest_unit_change_graph
```

```{r}
highest_pop_change_pivoted <- highest_pop_change_graph %>%
  pivot_longer(!County, names_to = "year", values_to = "population") %>%
    mutate(population  = population / 1000)
highest_pop_change_pivoted
```

```{r}
#pivot table and divide unit count by thousands
highest_unit_change_pivoted <- highest_unit_change_graph %>%
  pivot_longer(!County, names_to = "year", values_to = "housing_units") %>%
  mutate(housing_units  = housing_units / 1000)
highest_unit_change_pivoted
```

# Line graph of population ove time 
Todo: Add a Title and colors... 
```{r}
ggplot(highest_pop_change_pivoted, aes(x = year, y = population, group = County))  + 
  geom_line(aes(color=County)) + 
  geom_point(aes(color=County)) + 
  theme_minimal() + 
  labs(
    title = "Fastest Growing Texas Counties by Population",
    x =  "Year", y = "Population (1000s)",
    subtitle = "US Census 2000-2020",
    #caption = "Data: US Census"
  )
```

```{r}
ggplot(highest_unit_change_pivoted, aes(x = year, y = housing_units, group = County))  + 
  geom_line(aes(color=County)) + 
  geom_point(aes(color=County)) + 
  theme_minimal() + 
  labs(
    title = "Fastest Growing Texas Counties by Housing Units",
    x =  "Year", y = "Housing Units (1000s)",
    subtitle = "US Census 2000-2020",
    #caption = "Data: US Census"
  )
```

