---
title: "Census_Data_pull"
output: html_document
date: "2025 Spring"
---
Script to pull Census Data and map 

```{r}
# Save census API key
API_Key = '83a719c9a57b7c2fedd093f9a0108fcf8bf552eb'

```

Pull data from the census website for population, boundaries, etc .

Try the tidy census package from the census API development instructions
https://walker-data.com/tidycensus/articles/basic-usage.html
#API for census data https://www.census.gov/data/developers/data-sets.html

#https://edits.nationalmap.gov/apps/gaz-domestic/public/search/names 
# texas is code 48 places look up through USGS 
 
```{r}
library(tidycensus)
library(tidyverse)
library(choroplethr)
library(choroplethrMaps)
library(ggplot2)
library(dplyr)
library(tmap)
library(sf)
```

Get full list of variables in each year.
Full list of variables available at this URL : https://api.census.gov/data/2023/acs/acs5/profile/variables.html 
```{r}
acs13 <- load_variables(year = 2013,
                        dataset = "acs5", 
                        cache = TRUE)
```


```{r}
acs18 <- load_variables(year = 2018,
                        dataset = "acs5", 
                        cache = TRUE)
```

```{r}
acs23 <- load_variables(year = 2023,
                        dataset = "acs5", 
                        cache = TRUE)
```


```{r}
# Load variables from ACS 5-year profile
acs23_profile <- load_variables(year = 2023,
                        dataset = "acs5/profile", 
                        cache = TRUE)
```

Pull population for Texas 
The population was 29,640,343 in 2023 and 25,639,373 in 2013.

```{r}
tx_pop <- get_acs(geography = "state",
                     variables = c("DP05_0001"),
                     state = "TX",
                     survey = "acs5",
                     year = 2023)

tx_pop13 <- get_acs(geography = "state",
                     variables = c("DP05_0001"),
                     state = "TX",
                     survey = "acs5",
                     year = 2013)
```
The total number of housing units in 2023 was 11,890,808.
 In 2013, it was 10,070,703.
```{r}
tx_units <- get_acs(geography = "state",
                     variables = c("DP05_0091"),
                     state = "TX",
                     survey = "acs5",
                     year = 2023)


tx_units13 <- get_acs(geography = "state",
                     variables = c("DP04_0001"),
                     state = "TX",
                     survey = "acs5",
                     year = 2013)
```
```{r}
# median home value  2023 $260,400 
tx_value <- get_acs(geography = "state",
                     variables = c("B25077_001"),
                     state = "TX",
                     survey = "acs5",
                     year = 2023)

# median homevalue 2013 128900
tx_value13 <- get_acs(geography = "state",
                     variables = c("B25077_001"),
                     state = "TX",
                     survey = "acs5",
                     year = 2013)

value_change <- (tx_value$estimate - tx_value13$estimate )*100 / tx_value13$estimate
  
value_change
```


Pull ACS data from 2023
```{r}
# FETCH the data 
# DP05_0001 Total population 
# DP05_0091 number of housing units  # can confirm this is the same as total housing units (DP04)
# DP04_0001E total housing units in 2013 and 2023 but is depending on housing characteristics
# B25077_001 Median home value
# B07101_001E is the number of movers from another region... 

county_tx_pop <- get_acs(geography = "county",
                     variables = c("DP05_0001"),
                     state = "TX",
                     survey = "acs5",
                     year = 2023)

county_tx_hunit <- get_acs(geography = "county",
                 variables = c("DP05_0091"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2023)

county_tx_median_value <- get_acs(geography = "county",
                 variables = c("B25077_001"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2023)
# DP04_0001 2023
county_tx_hunit_occ <- get_acs(geography = "county",
                 variables = c("DP04_0001"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2023) 

#all NA values... hmmm? 
# county_tx_movers <- get_acs(geography = "county",
#                  variables = c("B07101_001"),
#                 state = "TX",
#               #  county = 201,
#                 survey = "acs5",
#                 year = 2023)
#
```

Let's look at 10 years ago
```{r}
county_tx_pop_2013 <- get_acs(geography = "county",
                     variables = "DP05_0001",
                     state = "TX",
                     survey = "acs5",
                     year = 2013)

county_tx_hunit_2013 <- get_acs(geography = "county", 
                 variables = c("DP04_0001"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2013)

county_tx_median_value_2013 <- get_acs(geography = "county",
                 variables = c("B25077_001"),
                state = "TX",
              #  county = 201,
                survey = "acs5",
                year = 2013)

```

Remove the variables we don't need and merge to create the dataset 
```{r}
county_tx_pop <- county_tx_pop %>% select(., c("GEOID", "NAME", estimate )) %>%
  rename(., population23 = estimate) 

county_tx_hunit <- county_tx_hunit %>% select(., c("GEOID", estimate )) %>%
  rename(., housing_unit23 = estimate) 
  
county_tx_median_value <- county_tx_median_value %>% select(., c("GEOID", estimate )) %>%
  rename(., median_value23 = estimate) 

county_tx_hunit_occ <- county_tx_hunit_occ %>% select(., c("GEOID", estimate )) %>%
  rename(., unit_occ23 = estimate) 

county_tx_pop_2013 <- county_tx_pop_2013 %>% select(., c("GEOID", estimate )) %>%
  rename(., population13 = estimate) 

county_tx_hunit_occ_2013 <- county_tx_hunit_2013 %>% select(., c("GEOID", estimate )) %>%
  rename(., unit_occ13 = estimate) 

county_tx_median_value_2013 <- county_tx_median_value_2013 %>% select(., c("GEOID", estimate)) %>%
  rename(., median_value13 = estimate) 


merged_census_tx <- county_tx_pop %>%
    left_join(., county_tx_hunit, 
            by = c("GEOID" = "GEOID")) %>% 
    left_join(., county_tx_median_value, 
            by = c("GEOID" = "GEOID")) %>% 
    left_join(., county_tx_hunit_occ, 
            by = c("GEOID" = "GEOID")) %>%
    left_join(., county_tx_pop_2013, 
            by = c("GEOID" = "GEOID")) %>%
    left_join(., county_tx_hunit_occ_2013, 
            by = c("GEOID" = "GEOID")) %>%
   left_join(., county_tx_median_value_2013, 
            by = c("GEOID" = "GEOID"))
  
```

```{r}
#save the dataset as a CSV 
write.csv(merged_census_tx, "~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/save_census_data_county.csv") 

summary(merged_census_tx)
```

TODO: #could also calculate a population or housing unit density.. 

Perform calculations to find the change in variables over 10 years
```{r}
#add new column from 2013 population estimate and also calcualte percentage change over the 10 years
#rename column 

merged_census_tx <- merged_census_tx %>%
  mutate(popchange10 = 
  (merged_census_tx$population23 - merged_census_tx$population13)*100 / merged_census_tx$population13) %>%
  mutate(unitchange10 = (unit_occ23 - unit_occ13)*100 / unit_occ13) %>%
   mutate(valuechange10 = (median_value23 - median_value13)*100 / median_value13) %>%
  
  merged_census_tx <- merged_census_tx %>% mutate(area = st_area(merged_census_tx$geometry))

  merged_census_tx <- merged_census_tx %>% mutate(pop_density = population23/st_area(merged_census_tx$geometry))

  


```



```{r}
pairs(merged_census_tx %>% select(3:12))
```

#visualize the population by county, shows there are very few with a big population... 
```{r}
merged_census_tx %>%
  ggplot(aes(y = population23, x = reorder(NAME, population23))) +
  geom_point()
```

Plot the growth over 10 years 
```{r}
merged_census_tx %>%
  ggplot(aes(y = popchange10, x = reorder(NAME, popchange10))) +
  geom_point()
```
Find five fastest growing counties since 2013

```{r}
#get the top values of tp_tx_county$tenyeargrowth 
#TODO: put the 2013 estimate on the dataframe... 
summary(merged_census_tx$popchange10)
#which values were these? 
tail(merged_texas[order(merged_texas$popchange10), ], 5)

```


Load in geography 

```{r}
# Get counties 
download.file("https://www2.census.gov/geo/tiger/TIGER2020PL/LAYER/COUNTY/2020/tl_2020_48_county20.zip", 
              destfile="Data/county_shape_48.zip")

county48 <-fs::dir_info(here::here("Data"  
                                 ))%>%  #$ means exact match
  dplyr::filter(str_detect(path, 
                           "tl_2020_48_county20.shp$"))%>%
  dplyr::select(path)%>%
  dplyr::pull()%>%
  
  sf::st_read() %>%
  st_transform(., 4326)

county48 <- st_read("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/tl_2020_48_county20.shp")

qtm(county48)
```
let's use ggplot2 to visualise... 


```{r}
# create a merged dataset with the geometry... 
texas_geo_subset <- county48 %>% select("GEOID20", "geometry")
merged_texas <- merged_census_tx %>%
  left_join(texas_geo_subset, by = c("GEOID" = "GEOID20"))




```

```{r}
ggplot(merged_texas) +
  geom_sf(aes(fill = popchange10, geometry = geometry)) +
    labs(
    title = "Population Change Across Texas Counties",
    subtitle = "Percent Change of 2013 to 2023",
    caption = "Data: American Community Survey"
  ) +
  theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1)
```
```{r}
ggplot(merged_texas) +
  geom_sf(aes(fill = unitchange10, geometry = geometry)) +
    labs(
    title = "Housing Unit Change Across Texas Counties",
    subtitle = "Percent Change of 2013 to 2023",
    caption = "Data: American Community Survey"
  ) +
  theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1) 
```
```{r}
#Questions: where did population grow and housing units did not?
#where did population grow and how did that compare to risk?

#which are the counties with the highest population change and highest risk score? 
highest_pop_change <- tail(merged_texas[order(merged_texas$popchange10), ], 5)
highest_unit_change <- tail(merged_texas[order(merged_texas$unitchange10), ], 5)
highest_value_change <- tail(merged_texas[order(merged_texas$valuechange10, na.last = FALSE), ], 5)

top_quintile_pop_change <- tail(merged_texas[order(merged_texas$popchange10), ], round(count(merged_texas) / 5)[[1]])

summary(merged_texas$popchange10)

highest_unit_change
```

```{r}
#map again 
ggplot() +
  geom_sf(data= merged_texas, aes(fill = popchange10, geometry = geometry)) +
    geom_sf(data= highest_pop_change, aes(color = "red",fill = popchange10, geometry = geometry)) +
    labs(
    title = "Population Change Across Texas Counties",
    subtitle = "Percent Change of 2013 to 2023, top five outlined in red",
    caption = "Data: American Community Survey"
  ) +
  theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1)
```


Regressions 
```{r}


#positive correlation between both the risk and the units... 
lm_fit <- lm(unitchange10 ~ RISK_SCORE, data=merged_texas)
summary(lm_fit)

lm_fit <- lm(popchange10 ~ RISK_SCORE, data=merged_texas)
summary(lm_fit)

lm_fit <- lm(valuechange10 ~ RISK_SCORE, data=merged_texas)
summary(lm_fit)

lm_fit <- lm(valuechange10 ~ popchange10, data=merged_texas)
summary(lm_fit)

ggplot() +  geom_point(data = merged_texas, aes(x =RISK_SCORE, y = unitchange10), alpha = 1 / 3) +
    geom_point(data = highest_unit_change, aes(x =  RISK_SCORE,  y= unitchange10), color = "Red" ) + 
  geom_smooth(data = merged_texas, aes(x =RISK_SCORE, y = unitchange10), method = "lm", se = FALSE)

#+ xlab("NRI Risk Score") + ylab("10 year Housing Unit change", )

ggplot() + geom_point(data = merged_texas, aes(x = RISK_SCORE, y= popchange10), alpha = 1 / 3) +
  #xlab("NRI Risk Score") + ylab("10 year population change") +  geom_point(alpha = 1 / 3) +
    geom_point(data = highest_pop_change, aes(x =  RISK_SCORE,  y= popchange10), color = "Red" ) + 
    geom_smooth(data = merged_texas, aes(x=RISK_SCORE, y = popchange10), method = "lm", se = FALSE) #+
  #   geom_label(data = highest_pop_change, aes(x =  RISK_SCORE,  y= popchange10), label = highest_pop_change$COUNTY, nudge_x=0.45, nudge_y=0.1, check_overlap=T)


ggplot() + geom_point(data = merged_texas, aes(x = RISK_SCORE, y= valuechange10), alpha = 1/3) + 
  #xlab("NRI Risk Score") + ylab("10 year median house value change") +  geom_point(alpha = 1 / 3)  +
    geom_point(data = highest_value_change, aes(x =  RISK_SCORE,  y= valuechange10), color = "Red" ) +
    geom_smooth(data = merged_texas, aes(x = RISK_SCORE, y= valuechange10), method = "lm", se = FALSE) #+
 #  geom_label(data = highest_value_change, aes(x =  RISK_SCORE,  y= valuechange10), label = highest_value_change$COUNTY, nudge_x=1, nudge_y=1, check_overlap=T)


```


```{r}
high_scoring <- merged_texas[(merged_texas$unitchange10 >40 & merged_texas$RISK_SCORE > 75 ),]


```

#TODO Wednesday: 
#something about adding the high scoring ones in highlight or with labels...? 
#then also add them with maps - highlight those counties
- label the known cities?


#identify coastal?
#also map the flooding? 

## Todo: add labels with the names of the cities... and then can we get the values? 
```{r}

p <- ggplot() +
  geom_sf(data = merged_texas, aes(fill = RISK_SCORE, geometry = geometry)) + 
  #geom_sf(data = texas_cities, aes(fill = "black", geometry = geometry)) 
#    geom_sf_text(data = texas_cities, aes(label = CITY_NAME, geometry = geometry), position = position_nudge(x = 0.6, y = 0.2)) +
      geom_sf(data= top_quintile_pop_change, aes(color = "red",fill = RISK_SCORE, geometry = geometry)) +
labs(
    title = "Risk Score Across Texas Counties",
    subtitle = "Climate Event Risk Index March 2023, Counties in top Quintile of population growth in red",
    caption = "Data: FEMA National Risk Index | ESRI"
  ) + 
  theme(
    plot.title = element_text(size = 24, hjust = 0.5)
  ) +
    theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1) 


```

```{r}
p
```
Autocorrelation for the whole state of texas and the counties... 
```{r}
library(spdep)

coordsW <- merged_texas$geometry%>%
  st_centroid()%>%
  st_geometry()
  
plot(coordsW,axes=TRUE)


#create a neighbours list
LWard_nb <- merged_texas$geometry %>%
  poly2nb(., queen=T)

summary(LWard_nb)


#plot them
plot(LWard_nb, st_geometry(coordsW), col="red")
#add a map underneath
plot(merged_texas$geometry, add=T)

#create a spatial weights matrix from these weights
Lward.lw <- LWard_nb %>%
  nb2mat(., style="B")

sum(Lward.lw)

sum(Lward.lw[1,])

Lward.lw <- LWard_nb %>%
  nb2listw(., style="C")

I_LWard_Global_Density <- merged_texas %>%
  pull(RISK_SCORE) %>%
  as.vector()%>%
  moran.test(., Lward.lw)

I_LWard_Global_Density


G_LWard_Global_Density <- 
  merged_texas %>%
  pull(RISK_SCORE) %>%
  as.vector()%>%
  globalG.test(., Lward.lw)

G_LWard_Global_Density
```
```{r}
Gi_LWard_Local_Density <- merged_texas %>%
  pull(RISK_SCORE) %>%
  as.vector()%>%
  localG(., Lward.lw)

head(Gi_LWard_Local_Density)

merged_texas <- merged_texas %>%
  mutate(density_G = as.numeric(Gi_LWard_Local_Density))
```

#Map the GI autocorrelation

```{r}
library(RColorBrewer)

GIColours<- rev(brewer.pal(8, "RdBu"))

breaks1<-c(-1000,-2.58,-1.96,-1.65,1.65,1.96,2.58,1000)


#now plot on an interactive map
# tm_shape(merged_texas$geometry) +
#     tm_polygons(col = merged_texas$density_G,
#         style="fixed",
#         breaks=breaks1,
#         palette=GIColours,
#         midpoint=NA,
#         title="Gi*, NRI texas")


ggplot() +
  geom_sf(data = merged_texas, aes(fill = density_G, geometry = geometry)) +
  labs(
    title = "Risk Score Autocorrelation Across Texas Counties",
    subtitle = "Climate Event Risk Index March 2023",
    caption = "Data: FEMA National Risk Index | ESRI"
  ) + 
  theme(
    plot.title = element_text(size = 24, hjust = 0.5)
  ) +
    theme_classic() +
  scale_fill_distiller(palette="YlGn" , direction = 1) 
```


<!-- Let's map Harris county (Aka Houston ) -->
<!-- ```{r} -->
<!-- #Get the tracts first  -->
<!-- download.file("https://www2.census.gov/geo/tiger/TIGER2020PL/LAYER/TRACT/2020/tl_2020_48_tract20.zip",  -->
<!--               destfile="Data/tract_shape_48.zip") -->

<!-- # texas_tracts <-fs::dir_info(here::here("Data"   -->
<!-- #                                  ))%>%  #$ means exact match -->
<!-- #   dplyr::filter(str_detect(path,  -->
<!-- #                            "tract_shape_48/tl_2020_48_tract20.shp$"))%>% -->
<!-- # ##  dplyr::select(path)%>% -->
<!-- #   dplyr::pull()%>% -->
<!-- #   sf::st_read() -->

<!-- texas_tracts <- st_read("~/Documents/_UCL_grad school/UCL/Dis/trial_project/Data/tract_shape_48/tl_2020_48_tract20.shp") %>% -->
<!--   st_transform(., 4326) -->

<!-- qtm(texas_tracts) -->


<!-- ``` -->


<!-- ```{r} -->
<!-- tp_hctx -->
<!-- ``` -->


<!-- ```{r} -->
<!-- print(mean(merged_texas$popchange10)) #1.47  -->
<!-- print(sd(merged_texas$popchange10)) # 16 -->
<!-- merged_texas$popchange10z <- (merged_texas$popchange10-mean(merged_texas$popchange10))/sd(merged_texas$popchange10) -->

<!-- ``` -->

```{r}
ggplot() +
  geom_sf(data = merged_texas, aes(fill = popchange10z, geometry = geometry)) +
#  geom_sf(data = texas_cities, aes(fill = "black", geometry = geometry )) +
  theme_void() +
  labs(
    title = "Population Change Across Texas Counties",
    subtitle = "Standard deviation of 2013 to 2023",
    caption = "Data: American Community Survey"
  ) + 
  theme(
    plot.title = element_text(size = 24, hjust = 0.5)
  )
```

```{r}
#TODO next: 
# map the climate indexes and the real estate prices
# try mapping only an area but down to the tract level (ie. Houston and Harriscounty or sth...)
  # then try doing some correlation? 

#Write up some introduction? 
```

Load in the National Risk INDEX (NRI)
```{r}
NRI_dictionary <- read_csv("~/Documents/_UCL_grad school/UCL/Dis/NRI_Table_Counties_Texas/NRIDataDictionary.csv")
```

```{r}
NRI_info <- read_csv("~/Documents/_UCL_grad school/UCL/Dis/NRI_Table_Counties_Texas/NRI_HazardInfo.csv")

NRI_tx <- read_csv("~/Documents/_UCL_grad school/UCL/Dis/NRI_Table_Counties_Texas/NRI_Table_Counties_Texas.csv")
```
#TODO: 
  Some additional questions: 
  - what is the distribution of the risk_score?  --> assuming its normalized to fit a 0-100 scale? 
  - how does population compare to the population from ACS? Very similar 
  - distribution of buildvalue? 
```{r}

NRI_tx_small <- NRI_tx %>% select("COUNTY", "COUNTYTYPE", "COUNTYFIPS", "STCOFIPS", "POPULATION", "BUILDVALUE",
                                  "AGRIVALUE", "RISK_VALUE", "RISK_SCORE", "EAL_VALB" ,"ALR_VALB") %>%
              mutate( STCOFIPS = as.character(STCOFIPS))
#could add additional variables about expected loss
```


```{r}
summary(NRI_tx_small)
print(mean(NRI_tx_small$RISK_VALUE))
```

```{r}
ggplot(NRI_tx_small, aes(x=RISK_VALUE)) + geom_histogram(binwidth=10000, colour="black", fill="white")


ggplot(NRI_tx_small, aes(x=RISK_SCORE)) + geom_histogram(binwidth=10, colour="black", fill="white")



```
Whats the distribution for the building loss rate? 
Percentage of total building value 
```{r}
ggplot(NRI_tx_small, aes(x=ALR_VALB)) + geom_histogram(binwidth=.00010, colour="black", fill="white")

```

```{r}
merged_texas <- merged_texas %>%
            left_join(NRI_tx_small, by = c("GEOID"= "STCOFIPS"))
       #      %>% select(., c("iso3", "country", "hdicode", "region",  "gii_2019", "gii_2020")) 
```

```{r}
world_cites <- st_read("https://services.arcgis.com/P3ePLMYs2RVChkJx/arcgis/rest/services/World_Cities/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson") %>%
  st_transform(., 4326)

texas_cities <- world_cites %>% filter(., FIPS_CNTRY == "US" & ADMIN_NAME == "Texas")

```

```{r}
q <- ggplot()+  geom_sf(data = texas_cities, aes(fill = "black", geometry = geometry )) + 
  geom_sf_text(data = texas_cities, aes(label = CITY_NAME, geometry = geometry), position = position_nudge(x = 0.6, y = 0.2)) +
 # geom_text(data = texas_cities, aes(label = CITY_NAME), size = 5) + 
  theme_void()
```

```{r}
q
```



Compare the population from NRI to population estimate 2023 from ACS 
#Tried and wasn't that helpful: graph the population 2013 versus 2023 

```{r}
ggplot() + geom_line(aes(x=merged_texas$POPULATION,y=merged_texas$popestimate2023),color='red') +
    ylab('ACS 2023 Population Estimate')+xlab('FEMA Population 2020')

```
